Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-09-23 Thesis/Background.tex
Final commit

Before proceeding to talk about the complications of executing TreePM on distributed systems like the Linux clusters worked with in this thesis, we would be remiss in not mentioning field expansion methods. Typically these take the form of an expansion of the potential in spherical harmonics using a set of radial basis functions. A brief discussion of these methods is presented in Chapter~2.9.4 of \citet{BT}. These methods have great historical significance, and have even seen a recent resurgence in helping understand the structure of dark matter halos \citep{lilley_2018_a, lilley_2018_b}.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../figures/Peanocurve.eps}
	\caption{Several examples of the Peano space filling curve in 2D. The level of refinement increases left-to-right. Obtained from \citet{peano}.}\label{fig:peano}
\end{figure}

@@

In this spirit, \citet{jeans_1915}\footnote{Notably before the famous 1920 Curtis-Shapley debate \citep{curtis_shapley} establishing that so-called ``spiral nebulae" were actually external galaxies.} laid out a framework to describe galaxy evolution in terms of the \textit{moments} of the distribution function. Specifically, the idea is to multiply the CBE (Eq. \eqref{eq:cbe_cartesian}) by $\textbf{v}_i$ or $\textbf{v}_i \textbf{v}_j$ and integrate over the velocity part of phase space. These together yield the second-order Jeans equations \citep{BT},
\begin{eqnarray}
\deriv{\nu}{t} + \deriv{\nu \langle v_i \rangle}{x_i} &=& 0 \\
\nu \deriv{\langle v_j \rangle}{t} + \nu \langle v_i \rangle \deriv{\langle v_j \rangle}{x_i} &=& -\nu \deriv{\Phi}{x_j} - \deriv{\nu \sigma^2_{ij}}{x_i}\label{eq:jeans}
\end{eqnarray}
where $i,j \in (1,2,3)$, $\langle\cdot\rangle$ denotes expectation over $f$, $\mathbb{E}_f[\cdot]$, $\nu$ is the number density of particles, and $\sigma^2_{ij} = \langle v_i v_j \rangle - \langle v_i \rangle \langle v_j \rangle$ is the velocity dispersion tensor. In equilibrium for spherical systems, the second equation is often written as \citep{BT},
\begin{equation}
\deriv{\nu \langle v_r^2 \rangle}{r} + \nu \left(\deriv{\Phi}{r} + \frac{2 \langle v_r^2 \rangle - \langle v_\theta^2\rangle - \langle v_\phi^2 \rangle}{r} \right)
\end{equation}
where $\theta$ is the polar angle, $\phi$ is the azimuthal angle, and $r$ is the spherical radius\footnote{Note a slight abuse of notation. We said that $v_i$ has $i \in (1,2,3)$, but here we use symbols to represent specific coordinate systems. The point is that the indices have three unique values.}. Note that there is only one scalar equation. Symmetries in the velocity dispersion tensor for spherical systems require that the cross moments are zero \citep{BT}. For axisymmetric systems, the equilibrium Jeans equations are often written \citep{BT},
\begin{eqnarray}
\deriv{\nu \langle v_R^2 \rangle}{R} + \deriv{\nu \langle v_R v_z \rangle}{z} + \nu \left(\frac{\langle v_R^2 \rangle - \langle v_\phi^2 \rangle}{R} + \deriv{\Phi}{R} \right) &=& 0\\
\frac{1}{R} \deriv{R \nu \langle v_R v_z \rangle}{R} + \deriv{\nu \langle v_z^2 \rangle}{z}+ \nu \deriv{\Phi}{z} &=& 0\\
\frac{1}{R^2} \deriv{R^2 \nu \langle v_R v_\phi \rangle}{R} + \deriv{\nu \langle v_z v_\phi\rangle}{z} &=& 0
\end{eqnarray}
where $z$ is the Cartesian $z$, $R$ is the cylindrical radius, and $\phi$ is the polar angle. These equations look suspiciously like Euler's equations for fluid dynamics with a missing energy equation. \citet{jeans_1915} did not magically solve the statistical mechanical issues mentioned in \S\ref{sec:motivation}; given a number density and potential, we still have an unknown velocity dispersion tensor with six elements, and only 4 independent equations. In the simplified spherical case, we have three unknown elements and two independent equations.

@@

Despite these shortcomings, more information can be imposed on the system to make these equations applicable. \citet{hernquist_1993} presented one of the first successful applications of Jeans modelling to creating equilibrium galaxies with bulges, discs, and dark matter halos. In the case of the dark matter halo for a disc-halo system, one only has to specify the function,
\begin{equation}
\beta(r) = 1 - \frac{\langle v_\theta^2\rangle + \langle v_\phi^2 \rangle}{2 \langle v_R^2 \rangle}.
\end{equation}
This parameter measures anisotropy in the velocity ellipsoid defined by $\sigma_{ij}$ at each radius; $\beta = -\infty$ if all orbits are circular, $\beta=0$ if the system is isotropic, and $\beta = 1$ if orbits are purely radial. Together with an assumed density-potential pair, continuity equation, and the spherical second-order Jeans equation, the spherical halo's velocity structure is specified. At the time of writing, \citet{hernquist_1993} did not have much information about the dark matter distribution in galaxies (recall \citep{nfw} followed this work in establishing a common density profile for dark matter halos). As such, we will not talk about their uninformed choice of halo, and it really is not important.

@@

Although there are many techniques that one can use to find clusters in data, we focus on \textit{hierarchical clustering} methods, which as the name suggests, are meant to build hierarchies of clusters. Since $\Lambda$CDM simulations produce subhalos within halos, we expect a such a structure in these clusters.

@@

\begin{figure}[h]
	\includegraphics[width=\textwidth]{../figures/eagle_tree}
	\caption{A merger tree obtained from \citet{merger_tree}. It represents the merging history of a halo in a simulation presented in \citet{Eagle}.} \label{fig:merger_tree}
\end{figure}

@@

Regardless of which approach to hierarchical clustering one uses, the end result is a tree-like structure called a \textit{dendrogram}, with large clusters at the top and individual particles at the bottom. In simulations of cosmology, we construct these at every timestep to determine which halos merged with other halos. There is a long literature on this subject described in \citet{rockstar}, although we use their implementation, the \textsc{rockstar} code, in this thesis.

The main idea behind the algorithm implemented in \textsc{rockstar} is to first apply a divisive scheme to the 3D data to create more manageable subclusters that get piped to an agglomerative 6D scheme. A big issue in the literature has been finding distance metrics which work with 6D phase space information in an acceptable manner \citep{rockstar}. In addition to the naive hierarchical clustering algorithm, \textsc{rockstar} also removes unbound particles from clusters and implements a clever scheme to tell when halos have merged in a snapshot.

The resultant data structure after applying this algorithm sequentially to all snapshots is  a dendrogram of dendrograms at all later times. A node in the last snapshot represents the merging history of a single galaxy, like the dendrogram in Fig. \ref{fig:merger_tree}. We see smaller halos coming together to form a Milky Way mass object at redshift $z=0$. By applying this technique to the raw particle data, we can pick out dark matter halos in our simulations.

>>>>

Before proceeding to talk about the complications of executing TreePM on distributed systems like the Linux clusters worked with in this thesis, we would be remiss in not mentioning field expansion methods. Typically these take the form of an expansion of the potential in spherical harmonics using a set of radial basis functions. A brief discussion of these methods is presented in Chapter~2.9.4 of \citet{BT}. These methods have great historical significance, and have even seen a recent resurgence in helping understand the structure of dark matter haloes \citep{lilley_2018_a, lilley_2018_b}.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../figures/Peanocurve.eps}
	\caption{Several examples of the Peano space filling curve in 2D. The level of refinement increases left-to-right. Obtained from \citet{peano} and reproduced under the Creative Commons Attribution-Share Alike 3.0 Unported license.}\label{fig:peano}
\end{figure}

@@

In this spirit, \citet{jeans_1915}\footnote{Notably before the famous 1920 Curtis-Shapley debate \citep{curtis_shapley}, a climax of the ongoing debate over whether so-called ``spiral nebulae" were actually external galaxies. Hubble's distance measurements to variable stars in M31 a few years later resolved the debate finally.} laid out a framework to describe galaxy evolution in terms of the \textit{moments} of the distribution function. Specifically, the idea is to multiply the CBE (Eq. \eqref{eq:cbe_cartesian}) by each of $(1,\textbf{v}_i)$ and integrate over the velocity part of phase space. These together yield the second-order Jeans equations \citep{BT},
\begin{eqnarray}
\deriv{\nu}{t} + \deriv{\nu \langle v_i \rangle}{x_i} &=& 0 \\
\nu \deriv{\langle v_j \rangle}{t} + \nu \langle v_i \rangle \deriv{\langle v_j \rangle}{x_i} &=& -\nu \deriv{\Phi}{x_j} - \deriv{\nu \sigma^2_{ij}}{x_i}\label{eq:jeans}
\end{eqnarray}
where $i,j \in (1,2,3)$, $\langle\cdot\rangle$ denotes expectation over $f$, $\mathbb{E}_f[\cdot]$, $\nu$ is the number density of particles, and $\sigma^2_{ij} = \langle v_i v_j \rangle - \langle v_i \rangle \langle v_j \rangle$ is the velocity dispersion tensor. In equilibrium for spherical systems, the second equation is often written as \citep{BT},
\begin{equation}
\deriv{\nu \langle v_r^2 \rangle}{r} + \nu \left(\deriv{\Phi}{r} + \frac{2 \langle v_r^2 \rangle - \langle v_\theta^2\rangle - \langle v_\phi^2 \rangle}{r} \right)
\end{equation}
where $\theta$ is the polar angle, $\phi$ is the azimuthal angle, and $r$ is the spherical radius\footnote{Note a slight abuse of notation. We said that $v_i$ has $i \in (1,2,3)$, but here we use symbols to represent specific coordinate systems. The point is that the indices have three unique values.}. Note that there is only one scalar equation. Symmetries in the velocity dispersion tensor for spherical systems require that the cross moments are zero \citep{BT}. For axisymmetric systems, the equilibrium Jeans equations are often written \citep{BT},
\begin{eqnarray}
\deriv{\nu \langle v_R^2 \rangle}{R} + \deriv{\nu \langle v_R v_z \rangle}{z} + \nu \left(\frac{\langle v_R^2 \rangle - \langle v_\phi^2 \rangle}{R} + \deriv{\Phi}{R} \right) &=& 0\\
\frac{1}{R} \deriv{R \nu \langle v_R v_z \rangle}{R} + \deriv{\nu \langle v_z^2 \rangle}{z}+ \nu \deriv{\Phi}{z} &=& 0\\
\frac{1}{R^2} \deriv{R^2 \nu \langle v_R v_\phi \rangle}{R} + \deriv{\nu \langle v_z v_\phi\rangle}{z} &=& 0
\end{eqnarray}
where $z$ is the Cartesian $z$, $R$ is the cylindrical radius, and $\phi$ is the polar angle. These equations look suspiciously like Euler's equations for fluid dynamics with a missing energy equation. \citet{jeans_1915} did not magically solve the statistical mechanical issues mentioned in \S\ref{sec:motivation}; given a number density and potential, we still have an unknown velocity dispersion tensor with six elements, and only 4 independent equations. In the simplified spherical case, we have three unknown elements and two independent equations.

@@

Despite these shortcomings, more information can be imposed on the system to make these equations applicable. \citet{hernquist_1993} presented one of the first successful applications of Jeans modelling to creating equilibrium galaxies with bulges, discs, and dark matter haloes. In the case of the dark matter halo for a disc-halo system, one only has to specify the function,
\begin{equation}
\beta(r) = 1 - \frac{\langle v_\theta^2\rangle + \langle v_\phi^2 \rangle}{2 \langle v_R^2 \rangle}.
\end{equation}
This parameter measures anisotropy in the velocity ellipsoid defined by $\sigma_{ij}$ at each radius; $\beta = -\infty$ if all orbits are circular, $\beta=0$ if the system is isotropic, and $\beta = 1$ if orbits are purely radial. Together with an assumed density-potential pair, continuity equation, and the spherical second-order Jeans equation, the spherical halo's velocity structure is specified. At the time of writing, \citet{hernquist_1993} did not have much information about the dark matter distribution in galaxies (recall \citep{nfw} followed this work in establishing a common density profile for dark matter haloes). As such, we will not talk about their uninformed choice of halo, and it really is not important.

@@

Although there are many techniques that one can use to find clusters in data, we focus on \textit{hierarchical clustering} methods, which as the name suggests, are meant to build hierarchies of clusters. Since $\Lambda$CDM simulations produce subhaloes within haloes, we expect a such a structure in these clusters.

@@

@@

Regardless of which approach to hierarchical clustering one uses, the end result is a tree-like structure called a \textit{dendrogram}, with large clusters at the top and individual particles at the bottom. In simulations of cosmology, we construct these at every timestep to determine which haloes merged with other haloes. There is a long literature on this subject described in \citet{rockstar}, although we use their implementation, the \textsc{rockstar} code, in this thesis.

The main idea behind the algorithm implemented in \textsc{rockstar} is to first apply a divisive scheme to the 3D data to create more manageable subclusters that get piped to an agglomerative 6D scheme. A big issue in the literature has been finding distance metrics which work with 6D phase space information in an acceptable manner \citep{rockstar}. In addition to the naive hierarchical clustering algorithm, \textsc{rockstar} also removes unbound particles from clusters and implements a clever scheme to tell when haloes have merged in a snapshot.

The resultant data structure after applying this algorithm sequentially to all snapshots is  a dendrogram of dendrograms at all later times. A node in the last snapshot represents the merging history of a single galaxy. In such a data structure, we see smaller haloes coming together to form a Milky Way mass object at redshift $z=0$. By applying this technique to the raw particle data, we can pick out dark matter haloes in our simulations.

### 2019-09-23 Thesis/Conclusion.tex
Final commit

We manage to bolster this result in Chapter~\ref{ch:paper_iii}, where we ran twelve simulations in four halos. Although all of the halos had the same total mass as the Milky Way, they varied widely in terms of their other properties. Despite the differences in the halos, the bars formed by embedded discs were quite similar. Additionally, there is a nontrivial dependence on Toomre $Q$ in terms of when the bar buckles, but the final bar strength depends very weakly on this parameter, contrary to \citet{YurinSpringelStellarDisks}. Rather, the trend \citet{YurinSpringelStellarDisks} sees in $Q$ is likely due to the presence of their bulge simulations, which would increase $Q$ as well as provide a stabilizing central mass.

@@

The main focus of Chapter~\ref{ch:paper_iii} is how discs form vertical structure in $\Lambda$CDM environments. We showed that there are a wide variety of Milky Way-mass halos that form vertical structure consistent with structure observed in the solar neighborhood. In particular, we showed that this structure can be formed without a massive satellite encounter in a cosmological setting. This position is bolstered by the fact that we see Monoceros-like structure in all twelve of our simulations.

@@

This work presents an improved disc insertion scheme with applications to Galactic astronomy. We have successfully shown that we can set up and understand discs inserted into cosmological halos. Future work is suggested, and we have no shortage of potential projects to improve our scheme. The long-term goal of this work is to bridge the gap with \textit{ab initio} simulations, and we have presented a clear roadmap for doing so.

As simulation resolutions and computing power improve, we may be forced to revisit the fundamentals at the beginning of Chapter~\ref{ch:background}. Simulation codes may need to be redesigned to accommodate ever-increasing needs to relate them to observations. Our hope is that disc insertion schemes will be at the heart of these discussions as a valuable tool in understanding observational data.

>>>>

We manage to bolster this result in Chapter~\ref{ch:paper_iii}, where we ran twelve simulations in four haloes. Although all of the haloes had the same total mass as the Milky Way, they varied widely in terms of their other properties. Despite the differences in the haloes, the bars formed by embedded discs were quite similar. Additionally, there is a nontrivial dependence on Toomre $Q$ in terms of when the bar buckles, but the final bar strength depends very weakly on this parameter, contrary to \citet{YurinSpringelStellarDisks}. Rather, the trend \citet{YurinSpringelStellarDisks} sees in $Q$ is likely due to the presence of their bulge simulations, which would increase $Q$ as well as provide a stabilizing central mass.

@@

The main focus of Chapter~\ref{ch:paper_iii} is how discs form vertical structure in $\Lambda$CDM environments. We showed that there are a wide variety of Milky Way-mass haloes that form vertical structure consistent with structure observed in the solar neighborhood. In particular, we showed that this structure can be formed without a massive satellite encounter in a cosmological setting. This position is bolstered by the fact that we see Monoceros-like structure in all twelve of our simulations.

@@

This work presents an improved disc insertion scheme with applications to Galactic astronomy. We have successfully shown that we can set up and understand discs inserted into cosmological haloes. Future work is suggested, and we have no shortage of potential projects to improve our scheme. The long-term goal of this work is to bridge the gap with \textit{ab initio} simulations, and we have presented a clear roadmap for doing so.

A common theme in our work is that both nature and nurture affect key aspects of the evolution of galaxies, namely in respect to the formation of bars and vertical structure. Future work will need to address this more quantitatively, as we work back to understanding our own Milky Way. In particular, we believe that disc insertion schemes will play a vital role in understanding the recently released Gaia DR2.

Additionally, as simulation resolutions and computing power improve, we may be forced to revisit the fundamentals at the beginning of Chapter~\ref{ch:background}. Simulation codes may need to be redesigned to accommodate ever-increasing needs to relate them to observations. Our hope is that disc insertion schemes will be at the heart of these discussions as a valuable tool in understanding observational data.
